---
title: "Engle Road - Road data from Twitter"
author: "Winston Saunders"
date: "November 8, 2014"
output: html_document
---

###Crash data analysis based on the TripCheck Twitter Feed

This analysis is based on data from the Oregon Dept of Transportation (ODOT) TripCheck feed on twitter. The feed tracks road conditions and could be useful for correlating events.   

It uses a separate program to download the data from the feed (TwitterReader) available in my github repository, which stores the feed data in a local .csv file.


Current analysis is plots of a timeline of specific events and a histogram and density anlaysis. 

```{r "get_tweets", echo=FALSE, warning=FALSE}

##This program assumes you have already run the TwitterReader.R to download a feed timeline. 
##It reads the data in teh form of a .csv 

## Get the relevnat tweet data

        if (getwd()=="/Users/winstonsaunders/Documents") {setwd("TripcheckR")}
        file_name <- "TripCheckUS20B"

        hwy_df<-read.csv(paste0(file_name, ".csv"))

        ##make the date column a date
        hwy_df$created<-as.Date(hwy_df$created)


```

The data pulled is from `r hwy_df$created[length(hwy_df$created)]` to `r hwy_df$created[1]`. There are `r dim(hwy_df)[1]` raw data points. 


```{r "clean", echo=FALSE, warning=FALSE}

        ##Define Search Patterns
        location <- "@ Engle Road"
        ##Get tweets with S1
        hwy_df <-hwy_df[grep(location, hwy_df$text),]
        
        incident <- "crash"
        ## get tweets with S2
        hwy_df <-hwy_df[grep(incident, hwy_df$text),]

##dedup the data

hwy_df<-hwy_df[!duplicated(hwy_df$created),]

        ##analyze the data

        ## get some details on the crash from the text string
        ##Find "Mi" maker and then extract miles number
        
        midist<-lapply(hwy_df$text, function(x) {y<- regexpr("Mi", x)
                                                             if (y >4) substring(x,y-3,y-1)
                                                             else "NA"})
                                                            # if (y != FALSE) substring(x, y-3, y-1)})
        
        distance<-as.numeric(midist)

        direction<-lapply(hwy_df$text, function(x) {y<- regexpr("Mi", x)
                                                             if (y >4) substring(x,y+2,y+3)
                                                             else "NA"})
        hwy_df$text<-substring(hwy_df$text,7,12)
        
        direction<-as.character(direction)

        ##bind new columns to data frame
        plotdata<-cbind(hwy_df, distance, direction )    

        ##reduce dimensions
        plotdata<-plotdata[,c(6,18,19)]

        ##filter for complete cases
        plotdata<-plotdata[complete.cases(plotdata), ]

        ##define function to strip spaces
        trim.leading <- function (x)  sub("^\\s+", "", x)

        plotdata$direction <- trim.leading(as.character(plotdata$direction))

        #Sign the distance based on direction

        for(i in 1:nrow(plotdata)) {if(plotdata$direction[i] =="E" | plotdata$direction[i] =="S") plotdata$distance[i] <- -1*plotdata$distance[i]}
                
                

##data are processed and ready to plot

```


Filtering on the crtieria: "`r location`"  and "`r incident`"  
Taking complete cases reduces the number of data points to `r dim(plotdata)[1]`.

```{r "plot_the_data", echo=FALSE, fig.align='center', fig.width=8, fig.height=4}


##timeline(plotdata)

        rangeYM<- c(plotdata$created[dim(plotdata)[1]], plotdata$created[1])

        ##index graph to start on first of month
        # first decompose into components
        d<-as.POSIXlt(rangeYM[1])
        e<-unclass(d)
        # Then subract appropriate number of days 
        rangeYM[1]<-rangeYM[1]-e$mday+1

```

The data cover a date range from `r rangeYM[1]` to `r rangeYM[2]` and there were `r 7 * nrow(plotdata)/as.numeric((rangeYM[2]- rangeYM[1]))` accidents per week.  

Since the distance over which the accidents occured is `r max(plotdata$distance)-min(plotdata$distance)` miles, the accident density is `r 52*7 * nrow(plotdata)/as.numeric((rangeYM[2]- rangeYM[1]))/(max(plotdata$distance)-min(plotdata$distance))` accident/year/mile. 

###Timline of accidents

The graph below shows a timline of accidents, with the location of the accident (measured in distance from the summit) represented by the y-axis. The positive direction is West.


```{r, echo=FALSE, fig.align='center'}

plot(NA,ylim=c(-10,10),xlim=rangeYM,ann=FALSE,axes=FALSE)
abline(h=0,lwd=2,col="dodgerblue4")

ypts <- plotdata$distance



y0pts = rep_len(0, length.out=nrow(plotdata))
txtpts <- rep_len(c(1,3), length.out=nrow(plotdata))
ylblpts <- ypts
lblpts <- rep_len(c(3,1), length.out=nrow(plotdata))
segments(as.Date(plotdata$created),y0pts,as.Date(plotdata$created),ypts,col="gray80")

axis.Date(
 1,
 at=seq.Date(rangeYM[1],rangeYM[2],by="month"),
 format="%Y-%m",
 cex.axis=0.6,
 pos=0,
 lwd=0,
 lwd.tick=2,
 col="dodgerblue3",
 font=1
)

points(plotdata$created,y=ypts, pch=20, cex=1.5, col="violetred2")

title(main=file_name, ylab="miles")



```



###Histogram of accident locations

This plot shows the distribution of accidents as measured by the distance from the point of reference `r location`. While the number of accidents on the West side outweigh those on the wast, there appears to be a strong bias to the distribution.  


```{r, echo=FALSE, fig.align='center'}


hist(plotdata$distance, main=file_name,xlab="distance (E-W)", col = "deepskyblue2", xlim=c(-10,10), breaks=10)
```

###Conclusion

This is a first analysis of this data. Will compare it to other road segments.