---
title: "Predicting Santiam Pass Accidents from Twitter"
author: "Winston Saunders"
date: "November 2014"
output: html_document
---

###Summary

The [Oregon Dept of Transportation](https://tripcheck.com/Pages/Twitter.asp) regularly publishes, as a public service, live road reports via Twitter. Using this data we can recontruct both the times and locations of accidents on sections of road and also correlate them with road conditions.  

The analysis focuses on a specific location, US Highway 20 at Santiam Pass, a 4800 foot (1450 meters) mountain pass in the Cascade Range (milepost 79). As the main route from Bend, Oregon to the Willamette Valley cities of Portland, Eugene, and Salem, it has high traffic year round and is the site of [frequent](http://www.nuggetnews.com/archives/960717/front1.shtml) accidents.

This analysis establishes the accident frequency and density for a 11 mile stretch of road to be correlation of snow and ice to traffic accidents on the pass. I show there is a high rate of accidents on the Pass (about one every two weeks). I use Bayesian analysis to show a high correlation to snowy road conditions. 


```{r "get_tweets", echo=FALSE, warning=FALSE}

##This program assumes you have already run the TwitterReader.R to download a feed timeline. 
##It reads the data in teh form of a .csv 

## Get the relevnat tweet data

        if (getwd()=="/Users/winstonsaunders/Documents") {setwd("TripcheckR")}
        file_name <- "TripCheckUS20B"

        hwy_df<-read.csv(paste0(file_name, ".csv"))

        ##make the date column a date
        hwy_df$created<-as.Date(hwy_df$created)

        ### define start adn edn times of period 
        tEnd <- hwy_df$created[length(hwy_df$created)]
        tStart <- hwy_df$created[1]
        


```

The data analyzed cover the dates from `r tEnd` to `r tStart`. There are `r dim(hwy_df)[1]` tweets during this period. 


```{r "clean", echo=FALSE, warning=FALSE}


        ##add meta table for correlation
        ## need to be insterted first
        metaincident <- "snow"
        ## get tweets with S2
        hwy_df2 <-hwy_df[grep(metaincident, hwy_df$text),]


        ##Define Search Patterns
        location <- "Santiam Pass Summit"
        ##Get tweets with S1
        hwy_df <-hwy_df[grep(location, hwy_df$text),]
        


        
        incident <- "crash"
        ## get tweets with S2
        hwy_df <-hwy_df[grep(incident, hwy_df$text),]
        
        

        

##dedup the data

        hwy_df<-hwy_df[!duplicated(hwy_df$created),]
        hwy_df2<-hwy_df2[!duplicated(hwy_df2$created),]

##analyze the hwy_df data

        ## get details on the crash from the text string

        ##Find "Mi" maker and then extract miles number
        
        midist<-lapply(hwy_df$text, function(x) {y<- regexpr("Mi", x)
                                                             if (y >4) substring(x,y-3,y-1)
                                                             else "NA"})
        
        distance<-as.numeric(midist)

        direction<-lapply(hwy_df$text, function(x) {y<- regexpr("Mi", x)
                                                             if (y >4) substring(x,y+2,y+3)
                                                             else "NA"})
        hwy_df$text<-substring(hwy_df$text,7,12)
        
        direction<-as.character(direction)

        ##bind new columns to data frame
        plotdata<-cbind(hwy_df, distance, direction )    

        ##reduce dimensions, keeping date, distance, and direction
        plotdata<-plotdata[,c(6,18,19)]

        ##filter for complete cases
        plotdata<-plotdata[complete.cases(plotdata), ]

        ##define function to strip spaces
        trim.leading <- function (x)  sub("^\\s+", "", x)

        plotdata$direction <- trim.leading(as.character(plotdata$direction))

        #Sign the distance based on direction

        for(i in 1:nrow(plotdata)) {if(plotdata$direction[i] =="E" | plotdata$direction[i] =="S") plotdata$distance[i] <- -1*plotdata$distance[i]}
                
           
## plot data 2

        plotdata2<-hwy_df2$created          ##keep only data column

##data are processed and ready to plot

        ##define a couple of variables for later use
        nAcc <- dim(plotdata)[1]
        nSnow <- length(plotdata2)

```


Filtering on the crtieria: _`r location`_  and _`r incident`_  and taking complete cases reduces the number of '`r incident` data points to `r nAcc`.   During the same period there were _`r nSnow`_ days with _`r metaincident`_. 

```{r "plot_the_data", echo=FALSE, fig.align='center', fig.width=8, fig.height=4}

##this section takes care of a formatting function to ensure the graph range starts on the first of a calendar month

##timeline(plotdata)

        rangeYM<- c(plotdata$created[dim(plotdata)[1]], plotdata$created[1])

        ##index graph to start on first of month
        # first decompose into components
        d<-as.POSIXlt(rangeYM[1])
        e<-unclass(d)
        # Then subract appropriate number of days 
        rangeYM[1]<-rangeYM[1]-e$mday+1

```

####summary statistics

There were `r 7 * nrow(plotdata)/as.numeric((rangeYM[2]- rangeYM[1]))` accidents per week. Since the distance over which the accidents occured is `r max(plotdata$distance)-min(plotdata$distance)` miles, the accident density is `r 52*7 * nrow(plotdata)/as.numeric((rangeYM[2]- rangeYM[1]))/(max(plotdata$distance)-min(plotdata$distance))` accident/year/mile. 

###Timeline of accidents

The graph below shows a timeline of accidents, with the location of the accident (measured in distance from the summit) represented by the y-axis. Red data points represent `r incident` data, while blue data points represent days when `r metaincident` was reported in the feed.


```{r, echo=FALSE, fig.align='center', fig.width=9  }

plot(NA,ylim=c(-10,10),xlim=rangeYM,ann=FALSE,axes=FALSE)

##manually draw axis and gridlines
abline(h=0,lwd=2,col="cornsilk4")
abline(h=2, lwd=1, col="cornsilk2")
abline(h=-2, lwd=1, col="cornsilk2")
abline(h=4, lwd=1, col="cornsilk2")
abline(h=-4, lwd=1, col="cornsilk2")
abline(h=6, lwd=1, col="cornsilk2")
abline(h=8, lwd=1, col="cornsilk2")

##define ypts as distance from summit
ypts <- plotdata$distance
##anchor line to x axis
y0pts2 <- rep_len(0, length.out=length(plotdata2))



y0pts <- rep_len(0, length.out=nrow(plotdata))
txtpts <- rep_len(c(1,3), length.out=nrow(plotdata))
ylblpts <- ypts
lblpts <- rep_len(c(3,1), length.out=nrow(plotdata))
segments(as.Date(plotdata$created),y0pts,as.Date(plotdata$created),ypts,col="cornsilk3")

##add y-axis numbers

axis(side=2, at=c(-4,-2,0,2,4,6,8), labels = c("-4 E", "-2 E", "0", "2 W", "4 W", "6 W", "8 W"), las=2, col="cornsilk4", lwd=2)

axis.Date(
 1,
 at=seq.Date(rangeYM[1],rangeYM[2],by="month"),
 format="%Y-%m",
 cex.axis=0.6,
 pos=0,
 lwd=0,
 lwd.tick=2,
 col="cornsilk4",
 font=1
)

points(plotdata$created,y=ypts, pch=20, cex=1.5, col="violetred2")
points(plotdata2, y = y0pts2, pch=19, cex=1, col="blue")

title(main=paste(location, file_name, "data", incident, "and", metaincident), ylab="distance from Summit (miles)")

##add x-axis text
text (rangeYM[1]+(rangeYM[2]-rangeYM[1])/2, -6, "date (yyyy-mm)")



```



###Histogram of distance from `r location`

This plot shows the distribution of accidents as measured by the distance from `r location`. In this specific case, while the number of accidents on the West side outweigh those on the East, there appears to be a strong bias to the distribution if one takes the mode as the center of the distribution.


```{r, echo=FALSE, fig.align='center'}


hist(plotdata$distance, main=file_name,xlab="distance (E-W)", col = "deepskyblue2", xlim=c(-10,10), breaks=10)
```


The median distance of accidents is `r median(plotdata$distance)` Miles West of the Summit with a standard deviation of `r sd(plotdata$distance)` about the mean `r mean(plotdata$distance)`. The data show a skewed distribution. 

###Probabilities

```{r, echo=FALSE}

tTotal <- as.numeric(tStart)-as.numeric(tEnd)

AccRate <- nAcc/tTotal
SnowRate<- nSnow/tTotal

```

We can calculate rough probabilities by noting that there were `r AccRate` crashes per day and the probability of snow was `r SnowRate` during the same period.


###Conclusion

This is a first analysis of this data. Will compare it to other road segments.