---
title: "TripCheckR - Road data from Twitter"
author: "Winston Saunders"
date: "October 26, 2014"
output: html_document
---

###Crash data anlaysis TripCheck Twitter Feed



```{r "get_tweets", echo=FALSE, warning=FALSE}

##This program assumes you have already run the TwitterReader.R to download a feed timeline. 
##It reads the data in teh form of a .csv 

## Get the relevnat tweet data

        if (getwd()=="/Users/winstonsaunders/Documents") {setwd("TripcheckR")}
        file_name <- "TripCheckUS20B"

        hwy_df<-read.csv(paste0(file_name, ".csv"))

        ##make the date column a date
        hwy_df$created<-as.Date(hwy_df$created)


```

The data pulled is from `r hwy_df$created[length(hwy_df$created)]` to `r hwy_df$created[1]`. There are `r dim(hwy_df)[1]` raw data points. 


```{r "clean", echo=FALSE, warning=FALSE}


        ##add meta table for correlation
        ## need to be insterted first
        metaincident <- "snow"
        ## get tweets with S2
        hwy_df2 <-hwy_df[grep(metaincident, hwy_df$text),]


        ##Define Search Patterns
        location <- "Santiam Pass Summit"
        ##Get tweets with S1
        hwy_df <-hwy_df[grep(location, hwy_df$text),]
        


        
        incident <- "crash"
        ## get tweets with S2
        hwy_df <-hwy_df[grep(incident, hwy_df$text),]
        
        

        

##dedup the data

        hwy_df<-hwy_df[!duplicated(hwy_df$created),]
        hwy_df2<-hwy_df2[!duplicated(hwy_df2$created),]

##analyze the hwy_df data

        ## get details on the crash from the text string

        ##Find "Mi" maker and then extract miles number
        
        midist<-lapply(hwy_df$text, function(x) {y<- regexpr("Mi", x)
                                                             if (y >4) substring(x,y-3,y-1)
                                                             else "NA"})
        
        distance<-as.numeric(midist)

        direction<-lapply(hwy_df$text, function(x) {y<- regexpr("Mi", x)
                                                             if (y >4) substring(x,y+2,y+3)
                                                             else "NA"})
        hwy_df$text<-substring(hwy_df$text,7,12)
        
        direction<-as.character(direction)

        ##bind new columns to data frame
        plotdata<-cbind(hwy_df, distance, direction )    

        ##reduce dimensions, keeping date, distance, and direction
        plotdata<-plotdata[,c(6,18,19)]

        ##filter for complete cases
        plotdata<-plotdata[complete.cases(plotdata), ]

        ##define function to strip spaces
        trim.leading <- function (x)  sub("^\\s+", "", x)

        plotdata$direction <- trim.leading(as.character(plotdata$direction))

        #Sign the distance based on direction

        for(i in 1:nrow(plotdata)) {if(plotdata$direction[i] =="E" | plotdata$direction[i] =="S") plotdata$distance[i] <- -1*plotdata$distance[i]}
                
           
## plot data 2

        plotdata2<-hwy_df2$created          ##keep only data column

##data are processed and ready to plot

```


Filtering on the crtieria: _`r location`_  and _`r incident`_ 
Taking complete cases reduces the number of data points to `r dim(plotdata)[1]`.   

The analysis also looks at meta incidents, in this case there were _`r length(plotdata2)`_ days with _`r metaincident`_. These data points are plotted along with the crash data.

```{r "plot_the_data", echo=FALSE, fig.align='center', fig.width=8, fig.height=4}

##this section takes care of a formatting function to ensure the graph range starts on the first of a calendar month

##timeline(plotdata)

        rangeYM<- c(plotdata$created[dim(plotdata)[1]], plotdata$created[1])

        ##index graph to start on first of month
        # first decompose into components
        d<-as.POSIXlt(rangeYM[1])
        e<-unclass(d)
        # Then subract appropriate number of days 
        rangeYM[1]<-rangeYM[1]-e$mday+1

```

The data cover a date range from `r rangeYM[1]` to `r rangeYM[2]` and there were `r 7 * nrow(plotdata)/as.numeric((rangeYM[2]- rangeYM[1]))` accidents per week.  

Since the distance over which the accidents occured is `r max(plotdata$distance)-min(plotdata$distance)` miles, the accident density is `r 52*7 * nrow(plotdata)/as.numeric((rangeYM[2]- rangeYM[1]))/(max(plotdata$distance)-min(plotdata$distance))` accident/year/mile. 

###Timline of accidents

The graph below shows a timline of accidents, with the location of the accident (measured in distance from the summit) represented by the y-axis. The positive direction is West.  
  

Blue data points represent days when `r metaincident` was reported in the feed.


```{r, echo=FALSE, fig.align='center', fig.width=9  }

plot(NA,ylim=c(-10,10),xlim=rangeYM,ann=FALSE,axes=FALSE)
abline(h=0,lwd=2,col="dodgerblue4")

ypts <- plotdata$distance

y0pts2 <- rep_len(0, length.out=length(plotdata2))



y0pts <- rep_len(0, length.out=nrow(plotdata))
txtpts <- rep_len(c(1,3), length.out=nrow(plotdata))
ylblpts <- ypts
lblpts <- rep_len(c(3,1), length.out=nrow(plotdata))
segments(as.Date(plotdata$created),y0pts,as.Date(plotdata$created),ypts,col="gray80")

axis.Date(
 1,
 at=seq.Date(rangeYM[1],rangeYM[2],by="month"),
 format="%Y-%m",
 cex.axis=0.6,
 pos=0,
 lwd=0,
 lwd.tick=2,
 col="dodgerblue3",
 font=1
)

points(plotdata$created,y=ypts, pch=20, cex=1.5, col="violetred2")
points(plotdata2, y = y0pts2, pch=19, cex=1, col="blue")

title(main=file_name, ylab="miles")



```



###Histogram of accident locations

This plot shows the distribution of accidents as measured by the distance from the point of reference `r location`. While the number of accidents on the West side outweigh those on the wast, there appears to be a strong bias to the distribution.  


```{r, echo=FALSE, fig.align='center'}


hist(plotdata$distance, main=file_name,xlab="distance (E-W)", col = "deepskyblue2", xlim=c(-10,10), breaks=10)
```

###Conclusion

This is a first analysis of this data. Will compare it to other road segments.